Model Recognition as Planning

### 2018-10-09 doc/icaps-19/model-recognition.tex
new full version icaps19 paper. Ready for reviews

\def\year{2017}\relax
%File: formatting-instruction.tex
\documentclass[letterpaper]{article} %DO NOT CHANGE THIS
\usepackage{aaai18}  %Required
\usepackage{times}  %Required
\usepackage{helvet}  %Required
\usepackage{courier}  %Required
\usepackage{url}  %Required
\usepackage{graphicx}  %Required
\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:
  \pdfinfo{}
\setcounter{secnumdepth}{0}

@@

\maketitle
\begin{abstract} 
Given the partial observation of a plan execution and a set of possible planning models (models that share the same state variables but update these variables according to different action models), {\em model recognition} is the task of identifying which model in the set explains (produced) the given observation. The paper formalizes the {\em model recognition} task and proposes a novel method to estimate the probability of a \strips\ model to produce a partial observation of a plan execution. This method, that we called {\em model recognition as planning}, is built on top of off-the-shelf classical planning algorithms and is robust to missing intermediate states and actions in the observed plan execution. The effectiveness of {\em model recognition as planning} is shown in a set of \strips\ models encoding different kinds of {\em automata}. We show that {\em model recognition as planning} succeeds to identify the executed automata despite the internal machine state or actual applied transitions, are unobserved.
\end{abstract}

\section{Introduction}
\label{sec:introduction}
{\em Plan recognition} is the task of predicting the future actions of an agent provided observations of its current behavior~\cite{carberry2001techniques}. {\em Goal recognition} is a closely related task that aims identifying the goals of the observed agent. Plan/goal recognition is considered {\em automated planning} in reverse; while automated planning compute sequences of actions that accounts for a given goals, plan/goal recognition compute goals that account for an observed sequence of actions~\cite{geffner:book:2013}.

Diverse approaches has been proposed for plan recognition such as {\em rule-based systems}, {\em parsing}, {\em graph-covering}, {\em Bayesian nets}, etc~\cite{sukthankar2014plan}. {\em Plan recognition as planning} is the model-based approach for plan recognition~\cite{ramirez2012plan,ramirez2009plan}. This approach assumes that the action model of the observed agent is known and leverages it to compute the most likely goal of the agent, according to the observed plan execution.

This paper formalizes the {\em model recognition} task where the object to recognize is not a goal but the {\em planning model} that determines the behavior of the observed agent. Given a partial observation of a plan execution and a set of possible planning models (models that share the same state variables but update these variables with different action models), {\em model recognition} is the task of identifying the model in the set with the highest probability of explaining (producing) the given observation.

\begin{figure}
  \begin{scriptsize}    
  \begin{center}
  \begin{tikzpicture}[scale=.4]
          \begin{scope}
            \draw (0, 0) grid (5, 5);
            \node[anchor=center] at (3.5, 1.5) {R};
            \node[anchor=center] at (0.5, 0.5) {};
            \draw[thick,style=dotted] (0.5,0.5) -- (4.5,0.5);
            \draw[thick,style=dotted] (4.5,0.5) -- (4.5,1.5);
            \draw[thick,style=dotted,->] (4.5,1.5) -- (3.7,1.5);                        
          \end{scope}
        \end{tikzpicture}
\hspace*{1cm}        
	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,semithick]
	  \node[state] (A)              {$q_0$};
	  \node[state] (B) [right of=A] {$q_1$};          
	  \path
(A) edge [bend left, align=center] node {inc-y,\\dec-y} (B)
(B) edge [bend left, align=center] node {inc-y,\\dec-y} (A);
	\end{tikzpicture}
  \end{center}
  \end{scriptsize}  
 \caption{\small ({\em Left)} Robot navigation plan in a $5\times 5$ grid. {\em (Right)} 2-state automata to control that the robot only increments its x-coordinate at {\em even} rows (i.e. when $q_0$ holds).}
\label{fig:grid-example}
\end{figure}

To better illustrate {\em model recognition}, imagine a robot in a $n\times n$ grid whose navigation is determined by the \strips\ model of Figure~\ref{fig:model-example}. According to this model the robot could increment its {\em x-coordinate} at {\em even rows} (when {\tt\small $q0$} holds) and decrement it at the {\em odd rows} (when {\tt\small $q1$} holds). Apart from this particular navigation model, different action models could be defined within the same state variables (e.g. altering the way {\tt\small $q0$} and {\tt\small $q1$} are required and updated) and these models can determine different kinds of robot navigation. Given an observation of a plan execution, like the one illustrated at Figure~\ref{fig:grid-example}, {\em model recognition} would aim here to identify which navigation model produced that observation, despite key information is unobserved (e.g. the particular applied actions or the value of the state variables {\small\tt $q0$} and {\small\tt $q1$}).

@@

The paper introduces {\em model recognition as planning}; a novel method to estimate the probability of a given \strips\ model to produce an observed plan execution. Our method is built on top of off-the-shelf classical planning algorithms, that are used to elicit the likelihood of the observations given a candidate model. {\em Model recognition as planning} is robust to missing intermediate states and actions in the observed plan execution.

We evaluate the effectiveness of {\em model recognition as planning} with a set of \strips\ models that represent different {\em automata}. All of these {\em automata} are defined within the same state variables (same input {\em alphabet} and same {\em machine states}) but different {\em transition functions}. We show that {\em model recognition as planning} succeeds to identify the executed {\em automata} despite internal machine states or actual applied transitions are unobserved.

@@

\section{Background}
\label{sec:background}
This section formalizes the models for {\em classical planning} and for the {\em observation} of a plan execution.

@@

A {\em classical planning frame} is a $\tup{F,A}$ pair, where $F$ is a set of fluents and $A$ is a set of actions. The semantics of actions $a\in A$ are specified with two functions: $\rho(s,a)$ that denotes whether the action is {\em applicable} in a state $s$ and $\theta(s,a)$ that denotes the {\em successor state} that results of applying the action $a$ in a state $s$.

In this work we specify action semantics (the $\rho$ and $\theta$ functions) with the \strips\ model. With this regard, an action $a\in A$ is defined with:
\begin{itemize}
\item $\pre(a)\in\mathcal{L}(F)$, the {\em preconditions} of $a$, is the set of literals that must hold for the action $a\in A$ to be applicable.
\item $\eff^+(a)\in\mathcal{L}(F)$, the {\em positive effects} of $a$, is the set of literals that are true after the application of action $a\in A$.
\item $\eff^-(a)\in\mathcal{L}(F)$, the {\em negative effects} of $a$, is the set of literals that are false after the application of the action.
\end{itemize}
We say that an action $a\in A$ is {\em applicable} in a state $s$ iff $\pre(a)\subseteq s$. The result of applying $a$ in $s$ is the {\em successor state} denoted by $\theta(s,a)=\{s\setminus\eff^-(a))\cup\eff^+(a)\}$.

A {\em classical planning problem} is a tuple $P=\tup{F,A,I,G}$, where $I$ is an initial state and $G\in\mathcal{L}(F)$ is a goal condition. A {\em plan} $\pi$ for $P$ is an action sequence $\pi=\tup{a_1, \ldots, a_n}$ that induces the {\em state trajectory} $\tup{s_0, s_1, \ldots, s_n}$ such that $s_0=I$ and, for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. The {\em plan length} is denoted with $|\pi|=n$ . A plan $\pi$ {\em solves} $P$ iff $G\subseteq s_n$, i.e.,~if the goal condition is satisfied at the last state reached after following the application of the plan $\pi$ in the initial state $I$. A solution plan for $P$ is {\em optimal} if it has minimum length.

@@

\subsection{The observation model}
Given a classical planning problem $P=\tup{F,A,I,G}$ and a plan $\pi$ that solves $P$; the observation of the execution of  $\pi$ on $P$ is $\tau=\tup{a_1^o,s_1^o \ldots , a_n^o, s_m^o}$, an interleaved combination of {\small $1\leq n\leq |\pi|$} observed actions and {\small $1\leq m\leq |\pi|$} observed states such that:
\begin{itemize}
\item Observed actions are {\em consistent} with $\pi$~\cite{ramirez2009plan}. This means that the sequence of observed actions $\tup{a_1^o, \ldots, a_n^o}$ in $\tau$ is a sub-sequence of the solution plan $\pi$.
\item Observed states are a sub-sequence  of partial states {\em consistent} with the {\em state trajectory} $\tup{s_0, s_1, \ldots, s_n}$ traversed by $\pi$. 
\end{itemize}

On the one hand, the initial state $I$ is fully observed while the observed states in $\tau$ may be partial, i.e. the value of certain fluents in the intermediate states may be omitted (~$|s_i^o|\leq |F|$ for every $1\leq i\leq m$). On the other hand, the sequence of observed states $\tup{s_1^o, \ldots, s_m^o}$ in $\tau$ is the same sequence of states traversed by $\pi$ but certain states may also be omitted. Formally, $0\leq|s_i^o|$ for every $1\leq i\leq m$. This means that the transitions between two consecutive observed states in $\tau$ may require the execution of more than a single action ($\theta(s_i^o,\tup{a_1,\ldots,a_k})=s_{i+1}^o$, where ${\small k\geq 1}$ is unknown and unbound). Therefore we can conclude that having $\tau$ does not implies knowing the actual length of $\pi$.

\begin{definition}[$\Phi$-observation]
Given a subset of fluents $\Phi\subseteq F$ we say that $\tau$ is a $\Phi$-observation of the execution of $\pi$ on $P$ iff, for every ${\small 1\leq i\leq m}$, each observed state $s_i^o$ only contains fluents in $\Phi$.
\end{definition}

For instance, Figure~\ref{fig:grid-example} illustrates the six-state $\Phi$-observation \{{\tt\footnotesize<(xcoord 2)(ycoord 1)>, <(xcoord 3)(ycoord 1)>, <(xcoord 4)(ycoord 1)>, <(xcoord 5)(ycoord 1)>, <(xcoord 5)(ycoord 2)>, <(xcoord 4)(ycoord 2)>}\}, where $\Phi$ only contains fluents of the kind {\tt\small (xcoord ?v)} and {\tt\small (ycoord ?v)}. This means that, for each observed state, only the value of fluents {\tt\small (xcoord ?v)} and {\tt\small (ycoord ?v)} is known while the value of the remaining fluents {\tt\small (next ?v1 ?v2)}, {\tt\small (q0)} and {\tt\small (q1)}, is unknown.

\section{Model Recognition}
\label{sec:recognition}
The {\em model recognition} task is a tuple $\tup{P,M,\tau}$ where:
\begin{itemize}
\item $P=\tup{F,A,I,G}$ is a {\em classical planning problem} s.t. the semantics of actions $a\in A$ is unknown, i.e. the corresponding functions $\rho(s,a)$ and/or $\theta(s,a)$ are undefined.
\item $M=\{\mathcal{M}_1,\ldots,\mathcal{M}_m\}$ is a finite non-empty {\em set of models} for the actions in $A$ s.t., each model in $\mathcal{M}\in M$, defines different function pairs $\tup{\rho,\theta}$ over the state variables $F$.
\item $\tau$ is an {\em observation} of the execution of a solution plan $\pi$ for $P$.
\end{itemize}

{\em Model recognition} can be understood as a {\em classification task} where each class is represented with a different planning model $\mathcal{M}\in M$ and the observed plan execution $\tau$ is the single example to classify. Further, the planning model associated to a class acts as a {\em class prototype} that summarizes all the plan executions that could be synthesized with that model (all the examples that belong to that class).

We follow the {\em naive Bayes classifier} to assign a model $\mathcal{M}\in M$ to the given observation $\tau$ wrt the following expression, $argmax_{\mathcal{M}\in M} P(\tau|\mathcal{M})\times P(\mathcal{M})$. The hypotheses in {\em model recognition} are then about the possible action models $\mathcal{M}\in M$, while the observation $\tau$ of a plan execution represents the input observation. The {\em solution} to the {\em model recognition} task is the model $\mathcal{M}\in M$ (or subset of models in $M$) that maximizes the previous expression.

The $P(\mathcal{M})$ probability expresses if one model is known to be a priori more likely than the others. If this probability is not given as input it is reasonable to assume that {\em a priori} all models are equiprobable. The challenge in our formulation for {\em model recognition} is the definition of the $P(\tau|\mathcal{M})$ {\em likelihood} that express the probability of observing $\tau$ when $\mathcal{M}$ is the planning model.

\section{Recognition of \strips\ models}
\label{sec:asPlanning}
Here we analyze the particular instantiation of the {\em model recognition} task where the semantics of actions (i.e. $\rho$ and $\theta$ functions) are specified with \strips\ action schemas. We start formalizing the \strips\ schema and define the full space of possible \strips\ schema. Eventually, we introduce an {\em edit distance} for \strips\ schema to estimate the $P(\tau|\mathcal{M})$ likelihoods for classical planning models.

\subsection{Well-defined \strips\ action schema}
\strips\ action schema provide a compact representation for specifying classical planning models. For instance, Figure~\ref{fig:model-example} shows six \strips\ action schema, codded in PDDL, that determine robot navigation in $n\times n$ grids.

A \strips\ action schema $\xi$ is defined by a list of {\em parameters} $pars(\xi)$, and three list of predicates ($pre(\xi)$, $del(\xi)$ and $add(\xi)$) that shape the kind of fluents that can appear in the {\em preconditions}, {\em negative effects} and {\em positive effects} of the actions induced from that schema.

We say that two \strips\ schemes $\xi$ and $\xi'$ are {\em comparable} iff $pars(\xi)=pars(\xi')$, both share the same list of parameters. For instance, we claim that the six action schema of Figure~\ref{fig:model-example} are {\em comparable} while the {\small\tt stack(?v1,?v2)} and {\small\tt pickup(?v1)} schemes from the four opertator {\em blocksworld}~\cite{slaney2001blocks} are not. Last but not least, we say that two \strips\ models $\mathcal{M}$ and $\mathcal{M}'$ are {\em comparable} iff there exists a bijective function $\mathcal{M} \mapsto \mathcal{M}^*$ that maps every action schema $\xi\in\mathcal{M}$ to a comparable schema $\xi'\in\mathcal{M'}$ and vice versa.

@@

\subsection{Edit distances for \strips\ action models}
{\em Edit distances} are similarity metrics traditionally computed over {\em strings} or {\em graphs} and that has been proved successful for {\em pattern recognition}~\cite{masek1980faster,bunke1997relation}. In this work we formalize and compute edit distances that are referred to \strips\ planning models.

>>>>

\def\year{2017}\relax
%File: formatting-instruction.tex
\documentclass[letterpaper]{article} %DO NOT CHANGE THIS
\usepackage{aaai19}  %Required
\usepackage{times}  %Required
\usepackage{helvet}  %Required
\usepackage{courier}  %Required
\usepackage{url}  %Required
\usepackage{graphicx}  %Required
\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:
  \pdfinfo{}
\setcounter{secnumdepth}{0}

@@

\maketitle
\begin{abstract} 
Given the partial observation of a plan execution and a set of possible planning models (models that share the same state variables but that update these variables according to different action models), {\em model recognition} is the task of identifying which model in the set explains/produced the given observation. The paper formalizes the {\em model recognition} task and introduces a novel method to estimate the probability of a \strips\ model to produce an observation of a plan execution. This method, that we called {\em model recognition as planning}, is built on top of off-the-shelf classical planning algorithms and is robust to missing intermediate states and actions in the observed plan execution. The effectiveness of {\em model recognition as planning} is shown in a set of \strips\ models encoding different kinds of {\em automata}. We show that {\em model recognition as planning} succeeds to identify the executed automata despite the internal machine state or actual applied transitions, are unobserved.
\end{abstract}

\section{Introduction}
\label{sec:introduction}
{\em Plan recognition} is the task of predicting the future actions of an agent provided observations of its current behavior~\cite{carberry2001techniques}. {\em Goal recognition} is a closely related task that aims identifying the goals of the observed agent. Goal recognition is considered {\em automated planning} in reverse; while automated planning compute sequences of actions that accounts for a given goals, goal recognition compute goals that account for an observed sequence of actions~\cite{geffner:book:2013}.

Diverse approaches has been proposed for plan/goal recognition such as {\em rule-based systems}, {\em parsing}, {\em graph-covering}, {\em Bayesian nets}, etc~\cite{sukthankar2014plan}. {\em Plan recognition as planning} is the model-based approach for plan/goal recognition~\cite{ramirez2012plan,ramirez2009plan}. This approach assumes that the action model of the observed agent is known and leverages it to compute the most likely goal of the agent, according to the observed plan execution.

This paper formalizes the {\em model recognition} task where the object to recognize is not a goal but the {\em planning model} that determines the behavior of the observed agent. Given a partial observation of a plan execution and a set of possible planning models (models that share the same state variables but that update these variables with different action models), {\em model recognition} is the task of identifying the model in the set with the highest probability of producing/explaining the given observation.

\begin{figure}
  \begin{scriptsize}    
  \begin{center}
  \begin{tikzpicture}[scale=.4]
          \begin{scope}
            \draw (0, 0) grid (5, 5);
            \node[anchor=center] at (3.5, 1.5) {R};
            \node[anchor=center] at (0.5, 0.5) {};
            \draw[thick,style=dotted] (0.5,0.5) -- (4.5,0.5);
            \draw[thick,style=dotted] (4.5,0.5) -- (4.5,1.5);
            \draw[thick,style=dotted,->] (4.5,1.5) -- (3.7,1.5);                        
          \end{scope}
        \end{tikzpicture}
\hspace*{1cm}        
	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,semithick]
	  \node[state] (A)              {$q_0$};
	  \node[state] (B) [right of=A] {$q_1$};          
	  \path
(A) edge [bend left, align=center] node {inc-y,\\dec-y} (B)
(B) edge [bend left, align=center] node {inc-y,\\dec-y} (A);
	\end{tikzpicture}
  \end{center}
  \end{scriptsize}  
 \caption{\small ({\em Left)} Robot navigating a $5\times 5$ grid. {\em (Right)} Automata controling that the robot only increments its x-coordinate at {\em even} rows (when $q_0$ holds, given that actions {\tt inc-y} and {\tt dec-y} update the robot y-coordinate and the automata state).}
\label{fig:grid-example}
\end{figure}

To better illustrate {\em model recognition}, imagine a robot in a $n\times n$ grid whose navigation is determined by the \strips\ model of Figure~\ref{fig:model-example}. According to this model the robot could increment its {\em x-coordinate} at {\em even rows} (when {\tt\small $q0$} holds) and decrement it at the {\em odd rows} (when {\tt\small $q1$} holds). Apart from this particular navigation model, different action models could be defined within the same state variables (e.g. altering the way {\tt\small $q0$} and {\tt\small $q1$} are required and updated) and these models can determine different kinds of robot navigation. Given an observation of a plan execution (like the one illustrated at Figure~\ref{fig:grid-example}) {\em model recognition} would aim here to identify which navigation model produced/explains that observation, despite key information is unobserved (e.g. the particular applied actions or the value of the state variables {\small\tt $q0$} and {\small\tt $q1$}).

@@

The paper also introduces {\em model recognition as planning}; a novel method to estimate the probability of a given \strips\ model to produce an observed plan execution. Our method is built on top of off-the-shelf classical planning algorithms and is robust to missing intermediate states and actions in the observed plan execution.

We evaluate the effectiveness of {\em model recognition as planning} with sets of \strips\ models that represent different {\em automata}. All the {\em automata} in a set are defined within the same state variables (same input {\em alphabet} and same {\em machine states}) but different {\em transition functions}. We show that {\em model recognition as planning} succeeds to identify the executed {\em automata} despite internal machine states or actual applied transitions are unobserved.

@@

\section{Background}
\label{sec:background}
This section formalizes the models for {\em classical planning} and for the {\em observation} of the execution of a classical plan.

@@

A {\em classical planning frame} is a $\tup{F,A}$ pair, where $F$ is a set of {\em fluents} and $A$ is a set of {\em actions}. The semantics of actions $a\in A$ are specified with two functions: $\rho(s,a)$ that denotes whether the action is {\em applicable} in a state $s$ and $\theta(s,a)$ that denotes the {\em successor state} that results of applying the action $a$ in a state $s$. In this work we specify action semantics (the $\rho$ and $\theta$ functions) with the \strips\ model. With this regard, an action $a\in A$ is defined by:
\begin{itemize}
\item $\pre(a)\in\mathcal{L}(F)$, the {\em preconditions} of $a$, is the set of literals that must hold for the action $a\in A$ to be applicable.
\item $\eff^+(a)\in\mathcal{L}(F)$, the {\em positive effects} of $a$, is the set of literals that are true after the application of action $a\in A$.
\item $\eff^-(a)\in\mathcal{L}(F)$, the {\em negative effects} of $a$, is the set of literals that are false after the application of the action.
\end{itemize}
We say that an action $a\in A$ is {\em applicable} in a state $s$ iff $\pre(a)\subseteq s$. The result of applying $a$ in $s$ is the {\em successor state} denoted by $\theta(s,a)=\{s\setminus\eff^-(a))\cup\eff^+(a)\}$.

A {\em classical planning problem} is a tuple $P=\tup{F,A,I,G}$, where $I$ is an initial state and $G\in\mathcal{L}(F)$ is a goal condition. A {\em plan} $\pi$ for $P$ is an action sequence $\pi=\tup{a_1, \ldots, a_n}$ that induces the {\em trajectory} $\tau(\pi,s_0)=\tup{a_1, s_1, \ldots, a_n, s_n}$ such that $s_0=I$ and, for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. The {\em plan length} is denoted with $|\pi|=n$. A plan $\pi$ {\em solves} $P$ iff $G\subseteq s_n$, i.e.,~if the goal condition is satisfied at the last state reached after following the application of the plan $\pi$ in the initial state $I$. A solution plan for $P$ is {\em optimal} if it has minimum length.

@@

\subsection{The observation model}
Given a classical planning problem $P=\tup{F,A,I,G}$ and a plan $\pi$ that solves $P$; the observation of the execution of $\pi$ on $P$ is $\mathcal{O}(\pi,I)=\tup{a_1^o,s_1^o \ldots , a_l^o, s_m^o}$, an interleaved combination of {\small $1\leq l\leq |\pi|$} observed actions and {\small $1\leq m\leq |\pi|$} observed states such that:
\begin{itemize}
\item Observed actions are {\em consistent} with $\pi$~\cite{ramirez2009plan}. This means that $\tup{a_1^o, \ldots, a_l^o}$ is a sub-sequence of the solution plan $\pi$.
\item Observed states are a sub-sequence  of partial states {\em consistent} with the sequence of states $\tup{s_0, s_1, \ldots, s_n}$ traversed by $\pi$. 
\end{itemize}

On the one hand, the initial state $I$ is fully observed while the observed states in $\mathcal{O}(\pi,I)$ may be partial, i.e. the value of certain fluents in the intermediate states may be omitted (~$|s_i^o|\leq |F|$ for every $1\leq i\leq m$). On the other hand, the sequence of observed states $\tup{s_1^o, \ldots, s_m^o}$ in $\mathcal{O}(\pi,I)$ is the same sequence of states traversed by $\pi$ but certain states may also be omitted. Formally, $0\leq|s_i^o|$ for every $1\leq i\leq m$. This means that the transitions between two consecutive observed states in $\mathcal{O}(\pi,I)$ may require the execution of more than a single action ($\theta(s_i^o,\tup{a_1,\ldots,a_k})=s_{i+1}^o$, where ${\small k\geq 1}$ is unknown and unbound). Therefore we can conclude that having $\mathcal{O}(\pi,I)$ does not implies knowing the actual length of $\pi$.

\begin{definition}[$\Phi$-observation]
Given a subset of fluents $\Phi\subseteq F$ we say that $\mathcal{O}(\pi,I)$ is a $\Phi$-observation of the execution of $\pi$ on $P$ iff, for every ${\small 1\leq i\leq m}$, each observed state $s_i^o$ only contains fluents in $\Phi$.
\end{definition}

Figure~\ref{fig:grid-example} illustrates the six-state $\Phi$-observation \{{\tt\footnotesize<(xcoord 2)(ycoord 1)>, <(xcoord 3)(ycoord 1)>, <(xcoord 4)(ycoord 1)>, <(xcoord 5)(ycoord 1)>, <(xcoord 5)(ycoord 2)>, <(xcoord 4)(ycoord 2)>}\}, where $\Phi$ only contains fluents of the kind {\tt\small (xcoord ?v)} and {\tt\small (ycoord ?v)}. This means that, for each observed state, only the value of fluents {\tt\small (xcoord ?v)} and {\tt\small (ycoord ?v)} is known while the value of the remaining fluents, namely {\tt\small (next ?v1 ?v2)}, {\tt\small (q0)} and {\tt\small (q1)}, is unknown.

\section{Model Recognition}
\label{sec:recognition}
The {\em model recognition} task is a tuple $\tup{P,M,\mathcal{O}}$ where:
\begin{itemize}
\item $P=\tup{F,A[\cdot],I,G}$ is a {\em classical planning problem} s.t. the semantics of actions $a\in A[\cdot]$ is unknown, i.e. the corresponding functions $\rho(s,a)$ and/or $\theta(s,a)$ are undefined.
\item $M=\{\mathcal{M}_1,\ldots,\mathcal{M}_m\}$ is a finite non-empty {\em set of models} for the actions in $A$ s.t., each model in $\mathcal{M}\in M$, defines different function pairs $\tup{\rho,\theta}$ over the state variables $F$.
\item $\mathcal{O}(\pi,I)$ is an {\em observation} of the execution of an unknown solution plan $\pi$ for the planning problem $P$.
\end{itemize}

{\em Model recognition} can be understood as a {\em classification task} where each class is represented with a different planning model $\mathcal{M}\in M$ and the observed plan execution $\mathcal{O}(\pi,I)$ is the single example to classify. Further, the planning model associated to each class acts as the corresponding {\em class prototype} that summarizes all the plan executions that could be synthesized with that model (i.e. all the examples that belong to that class).

In this work we follow the {\em naive Bayes classifier} to assign a model $\mathcal{M}\in M$ to the given observation $\mathcal{O}(\pi,I)$ with respect to the expression:
\begin{align}
argmax_{\mathcal{M}\in M} P(\mathcal{O}|\mathcal{M})\times P(\mathcal{M}).
\end{align}
The hypotheses in {\em model recognition} are then about the possible action models $\mathcal{M}\in M$, while the observation $\mathcal{O}(\pi,I)$ of a plan execution represents the input observation. The {\em solution} to the {\em model recognition} task is the model $\mathcal{M}\in M$ (or subset of models in $M$) that maximizes the previous expression.

The $P(\mathcal{M})$ probability expresses whether one model is known to be a priori more likely than the others. If this probability is not given as input it is reasonable to assume that {\em a priori} all models are equiprobable. The challenge in our formulation for {\em model recognition} is the definition of the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} that expresses the probability of observing $\mathcal{O}(\pi,I)$ when $\mathcal{M}$ is the planning model.

\begin{figure}
  \begin{scriptsize}
  \begin{center}
	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,semithick]
	  \node[state] (A)              {$\mathcal{M}$};
	  \node[state] (B) [right of=A] {$\mathcal{M}'$};
	  \node[state] (C) [right of=B] {$\pi$};
	  \node[state] (D) [right of=C] {$\mathcal{O}$};          
	  \path
(A) edge  node {} (B)
(B) edge  node {} (C)
(C) edge  node {} (D)
;
	\end{tikzpicture}
  \end{center}
  \end{scriptsize}  
 \caption{\small {\em Bayesian network} abstracting that model $\mathcal{M}'$ results from editing $\mathcal{M}$ and that can produce a plan $\pi$ consistent with $\mathcal{O}(\pi,I)$.}
\label{fig:net}
\end{figure}

Our approach to formulate the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} is to assess how far is $\mathcal{M}$ from a model $\mathcal{M'}$ that can produce a plan $\pi$ that solves $P$ and s.t. $\mathcal{O}(\pi,I)$ is a {\em consistent} observation of the execution of $\pi$ on $P$. Figure~\ref{fig:net} shows the four-variable {\em Bayesian network} abstracting this process. Regarding this network we have that:
\begin{align}
 P(\mathcal{O}|\mathcal{M})=\sum_{\mathcal{M}'}\sum_{\pi} P(\mathcal{M'}|\mathcal{M})P(\pi|\mathcal{M'})P(\mathcal{O}|\pi),
\end{align}
where $\mathcal{M}'$ ranges over all the models that can be generated editing $\mathcal{M}$ and $\pi$ ranges over all the plans that can be synthesized with a model $\mathcal{M}'$

\subsection{Approximating the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood}}
Following the previous equation (2) the exact computation of $P(\mathcal{O}|\mathcal{M})$ is intractable because, for most planning problems, the set of {\em valid} plans consistent with an observation can easily be very large (infinite in the case of planning problems without dead-ends). Instead, in this work we propose to estimate $P(\mathcal{O}|\mathcal{M})$ making the following assumptions:
\begin{enumerate}
\item The sum in the previous equation is dominated by the {\em closest compliant set}, i.e. the models closest to $\mathcal{M}$ that can produce a solution plan consistent with $\mathcal{O}(\pi,I)$. The rationale behind this assumption is that the further the compliant model, the lower the $P(\mathcal{O}|\mathcal{M})$ likelihood. 
\end{enumerate}

The {\em full observability of the executed plan} is a too strong assumption but, as we explain next, it allows us to build reasonable estimates of the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} for the more general case where the observability of the executed plan is partial.

Note that in this baseline scenario where there is {\em full observability of the executed plan}:
\begin{itemize}
\item There is only a single possible plan {\em consistent} with the given observation, besides $P(\mathcal{O}|\pi)=1$.
\item If we consider assumption 1, the {\em closest compliant set} $\mathcal{M}^*$ becomes a singleton for this particular scenario, so that probabilities corresponding to different compliant models are not added up. Further, the $P(\pi|\mathcal{M^*})$ probability equals also to $1$ since there is only one possible plan consistent with the observation and by definition, the {\em closest compliant} model produces that plan.
\end{itemize}
As a consequence, when there is {\em full obsevability of the executed plan} and under the previous assumption, we have that expression (1) simplifies to
\begin{align}
argmax_{\mathcal{M}\in M} P(\mathcal{M^*}|\mathcal{M}).
\end{align}

Next we show how to approximate the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} for the more general case where there is partial observability of the executed plan. First we start with the simpler scenario where the length of the observed plan is bound and add these two new assumptions.
\begin{enumerate}\addtocounter{enumi}{1}
\item The {\em closest compliant set} is a singleton, so that probabilities corresponding to different compliant models are not added up.
\item All solutions of a given {\em plan length} are equiprobable, with shorter plans more likely than longer plans.  
\end{enumerate}

In such an scenario $P(\mathcal{O}|\pi)$ equals to a constant $\alpha$ that indicates the number of plans of bounded length that are consistent with $\mathcal{O}$. Likewise, $P(\pi|\mathcal{M^*})$ is also one given the definition of the {\em closest compliant set}. As a consequence, when there is {\em partial obsevability of the executed plan} (but the plan length is bound), and under the previous three assumptions, we have that expression (1) simplifies to
\begin{align}
argmax_{\mathcal{M}\in M} P(\mathcal{M^*}|\mathcal{M})\times \alpha.
\end{align}
where $\alpha$ is a normalizing constant that has the same value for every $\mathcal{M}\in M$, so it can be ignored.

Finally, we can say that the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} can be estimated for the general case of partial observability of the executed plan (where the length of the plan is not bound) by adding a fourth assumption:
\begin{enumerate}\addtocounter{enumi}{3}
\item The observed plan $\pi$ that solves $P$ is an optimal plan since we are assuming that the observed agent is acting rationally~\cite{ramirez2012plan}.
\end{enumerate}
In this case the length of the plan becomes bound by the optimal plan length and hence, we can again use expression (4) for the  likelihood estimation.

\section{Recognition of \strips\ models}
\label{sec:asPlanning}
Here we analyze the particular instantiation of the {\em model recognition} task where the semantics of actions (i.e. $\rho$ and $\theta$ functions) are specified with \strips\ action schemas. We start formalizing the \strips\ schema and define the full space of possible \strips\ schema. Eventually, we introduce an {\em edit distance} for \strips\ schema to estimate the $P(\mathcal{O}|\mathcal{M})$ likelihoods for classical planning models.

\subsection{Well-defined \strips\ action schema}
\strips\ action schema provide a compact representation for specifying classical planning models. For instance, Figure~\ref{fig:model-example} shows six \strips\ action schema, codded in PDDL, that determine a particular kind of robot navigation in $n\times n$ grids.

A \strips\ action schema $\xi$ is defined by a list of {\em parameters} $pars(\xi)$, and three list of predicates (namely $pre(\xi)$, $del(\xi)$ and $add(\xi)$) that shape the kind of fluents that can appear in the {\em preconditions}, {\em negative effects} and {\em positive effects} of the actions induced from that schema.

We say that two \strips\ schemes $\xi$ and $\xi'$ are {\em comparable} iff $pars(\xi)=pars(\xi')$, both share the same list of parameters. For instance, we claim that the six action schema of Figure~\ref{fig:model-example} are {\em comparable} while, for example, the {\small\tt stack(?v1,?v2)} and {\small\tt pickup(?v1)} schemes from the four opertator {\em blocksworld}~\cite{slaney2001blocks} are not. Last but not least, we say that two \strips\ models $\mathcal{M}$ and $\mathcal{M}'$ are {\em comparable} iff there exists a bijective function $\mathcal{M} \mapsto \mathcal{M}^*$ that maps every action schema $\xi\in\mathcal{M}$ to a comparable schema $\xi'\in\mathcal{M'}$ and vice versa.

@@

\subsection{Edit distances for \strips\ action models}
{\em Edit distances} are similarity metrics, traditionally computed over {\em strings} or {\em graphs}, and that has been proved successful for {\em pattern recognition}~\cite{masek1980faster,bunke1997relation}. In this work we formalize and compute edit distances that are referred to \strips\ planning models.  First, we define two edit \emph{operations} on a schema $\xi\in\mathcal{M}$ that belongs to a \strips\ action model $\mathcal{M}\in M$:
\begin{itemize}
\item {\em Deletion}. An element is removed from any of these three lists $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of the operator schema $\xi\in\mathcal{M}$ such that the resulting schema is a {\em well-defined} \strips\ action schema.
\item {\em Insertion}. An element in ${\mathcal I}_{\Psi,\xi}$ is added to any of these three lists $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of the operator schema $\xi\in\mathcal{M}$ s.t. the resulting schema is {\em well-defined}.
\end{itemize}
